Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-15 manuscript/Ydiversity_nov12.tex
edits

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plants and animals \citep{westergaard1958,ohno1967,bull1983,charlesworth1991,charlesworth2015plant}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes has been associated with an accumulation of deleterious mutations and a gradual loss of genes from the Y chromosome \citep{hough2014,bergero2015,bachtrog2013NRG}, and in some species, the degeneration of the Y has led to the evolution dosage compensation of the X chromosome \citep{charlesworth1996CB,muyle2012,mank2013sex,papadopulos2015}. The independent evolution of these phenomena in such taxonomically distant species suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in genomic regions that lack recombination is a longstanding challenge for both theoreticians and experimentalists \citep{charlesworth1978,feldman1980evolution,barton1995general,charlesworth1996CB,otto1997deleterious,charlesworth2000degeneration,mcvean2000effects}.

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Thus, genes on the Y chromosome are expected to experience an effective population size ($N_{e}$) that is 1/4 that of autosomal genes, whereas the  $N_{e}$ for genes on the X chromosome should be 3/4 that of autosomal genes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that that the equilibrium level of neutral polymorphism - proportional to the product of $N_{e}$ and the neutral mutation rate, $\mu$ - should be lower for Y-linked genes than for their X-linked counterparts. In the absence of recombination, genes on the Y chromosome are also expected to be in strong linkage disequilibrium, making them vulnerable to diversity loss due to selection against strongly deleterious mutations (background selection) and selective sweeps of strongly beneficial mutations (genetic hitchhiking). Furthermore, the build-up of linkage disequilibrium between selected mutations on the Y means that selection will act non-independently across the chromosome such that selection at a focal site may "interfere" with selection at sites with which it is linked. Originally developed by Hill & Robertson (1966), a large body of work has now shown that such "selective interference" can substantially reduce both the efficacy of selection and the level of neutral variability \citep{fisher1930genetical, muller1964relation, hill1966HReffect, mcvean2000}. These arguments all suggest that non-recombining Y chromosomes should harbor a lower amount of neutral genetic variability than predicted based on the number of Y chromosomes in a population, but the extent to which background selection, genetic hitchhiking, or selective interference have affected chromosome-wide levels of diversity in natural populations, and the relative importance of these processes, is not well-understood.

In addition to reduced diversity arising from selection, in species with female-biased sex ratios or extensive male-male competition, high variance in male reproductive success is also expected reduce the $N_{e}$ experience by genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting that inferences about the effects positive of purifying selection need to be distinguished from these neutral processes. Because variance in male reproductive success reduces both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this can therefore be obtained by comparing levels of neutral diversity on X and Y chromosomes relative to values on autosomes. For example, high variance in male reproductive success causes a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans, for example, have suggested that the inflated X/A ratio is attributable to a historical excess of breeding females over males \citep{hammer2008sex} (and see \citep{bustamante2009,hammer2010,cotter2016genetic}).

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on sex chromosomes \citep{ellegren2011,bachtrog2013NRG}, we know very little about the influence of either sex ratio variation or linked selection in determining levels of diversity on more recently evolved sex chromosomes. The time scales over which these different effects are likely to be important is therefore not well understood. In humans, estimates of Y-linked diversity are considerably lower than predicted under neutral models, and simulations suggest that levels of diversity are consistent with strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is not immediately obvious whether purifying selection might have such strong effects on Y chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA in the case of plants \citep{charlesworth2015plant}). Y-linked diversity loss might on the one hand be expected to be lower in younger systems due to a shorter history of recombination suppression. On the other hand, simulations of strong selection models (background selection and genetic hitchhiking) suggest that these processes may have the greatest effects during the earliest stages of sex chromosome evolution, before the Y has lost many of its genes \citep{bachtrog2008temporal}. Moreover, work by \citep{KaiserCharlesworth} and \citep{good2014genetic} have provided evidence that even weak purifying selection, if acting non-independently across large number of sites, can generate substantial deviations from neutrality, whereas classic background selection theory breaks down in such cases. Given that a large number of sites are likely to be in linkage disequilibrium on a recently evolved Y chromosome, such "interference selection" \textit{sensu} \citep{good2014genetic} is a good candidate model for exploring the evolutionary dynamics of young plant sex chromosomes.

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution, we analyzed neutral polymorphism levels on X, Y, and autosomal chromosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with heteromorphic X and Y chromosomes that evolved approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making Y chromosomes in this species over 100 million years younger than the highly degenerated Y chromosomes in mammals \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the occurrence of an intraspecific polymorphism in sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct populations ("chromosomal races") \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species is thought to have originated through an X-autosome fusion, with the XY system maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, there is evidence that both Y's have undergone significant gene loss and functional deterioration \citep{hough2014}. Here, to simplify our comparison of polymorphism levels on X, Y, and autosomes, we focus only on the ancestral Y chromosome, which occurs in both sex chromosome races.

Of particular importance for our study, \textit{R. hastatulus} populations have been found to consistently exhibit female-biased reproductive sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Female-biased sex ratios are not uncommon in dioecious plants with heteromorphic X and Y chromosomes \citep{field2013comparative,hough2013evolutionarily}, and their occurrence in \textit{R. hastatulus} provides an excellent opportunity to study both the demographic and selective factors contributing to sex-linked variability.

@@

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether our estimated levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, we compared our estimated levels diversity to predictions from a model of the expected $N_{e}$ of males and females when the variance in offpring number is either smaller or larger than Poisson \citep{kimura1964number, hedrick2011genetics}. As we were primarily interested in the possible influence of variance in male reproductive success on Y-linked $N_{e}$, and the parameter space within which such a model might generate Y/A and X/A ratios consistent with our data, we ignored non-poission variance in female offspring, and calculated the male $N_{e}$ as:

@@

where $k_{m}$ and $V_{k_{m}}$ are the mean and variance in male offspring number, respectively, in a population of $N_{m}$ reproducing males, and $\overline{k}$ is the total mean number of progeny \citep{kimura1964number}. Given expected values of $N_e_{m}$ for a range of values for $V_{k_{m}}$ and $N_{m}$, the corresponding expectation for the $N_{e}$ of autosomes and sex chromosomes at equilibrium were calculated as:

\begin{equation}
N_{e{A}} = \frac{4N_{m}N_{f}}{N_{m}+N_{f}} \label{eq:Ne}
\end{equation}

\begin{equation}
N_{e{X}} = \frac{9N_{m}N_{f}}{4N_{m}+2N_{f}} \label{eq:NeX}
\end{equation}

\begin{equation}
N_{e{Y}} = \frac{N_{m}}{2} \label{eq:NeY}
\end{equation}

following \citep{wright1931evolution}. Note that with equal sex ratios $N_{f} = N_{m}$, and Poisson-distributed offspring numbers ($V_{k} = \overline{k}$), then $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$. Figure 1 shows these predictions as a function of the sex ratio for a range values of $V_{k_{m}}$, and we tested the fit of our empirical diversity ratios to these predictions (shown in Figure 2) assuming the estimated \textit{R. hastatulus} sex ratio of 0.6 \citep{pickup2013influence}.

@@

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we used a similar approach to \citep{Wilsonsayres2014} and conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious mutations from our polymorphism data for X-linked genes using the method of \citep{keightley2007joint}, which fits a gamma distribution of selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphism. We used this estimated gamma distribution to parameterize the simulations, initializing each with our estimated $\theta$ from autosomal genes, but appropriately adjusted to reflect the expected reduction in Y-linked $N_{e}$ for a sex ratio of $N_{f}/(N_{f}+N_{m})=0.6$. To match our sample size and the number of synonymous sites sample from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and simulated sequences contained 45,331bp of linked neutral sequence from which we calculated diversity. To examine the expected reduction in diversity relative to neutrality ($\pi/\pi_{0}$) as a function of the number of selected sites ($L$), and to estimate this parameter from our data, we ran simulations over a range of values of $L$ (up to a maximum of $5x10^{6}$) (Figure 3A). For each simulation, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi_{s}_{simulated}$, matched our empirical estimate, $\pi_{s}_{observed}$ (Figure 3B).

\section*{Results and Discussion}

\subsection*{Extensive loss of Y-chromosome diversity}

\subsection*{Female biased sex ratio and variance in male offspring number}

\begin{figure}[htbp]
\centering
\noindent
\includegraphics[width=\linewidth]{figure2.jpg}
\caption{new figure coming soon).
}
\label{fig:spectrum}
\end{figure}

>>>>

\lettrine[lines=2]{\color{color2}M}{}orphologically distinct sex chromosomes have evolved multiple times independently in both plants and animals \citep{westergaard1958,ohno1967,bull1983,charlesworth1991,charlesworth2015plant}. Despite clear biological differences between these kingdoms, X and Y chromosomes in both lineages have undergone similar genetic changes. For example, in both groups the loss of recombination between X and Y chromosomes has been associated with an accumulation of deleterious mutations and a gradual loss of genes from the Y chromosome \citep{hough2014,bergero2015,bachtrog2013NRG}, and in some species, the degeneration of the Y has led to the evolution dosage compensation of the X chromosome \citep{charlesworth1996CB,muyle2012,mank2013sex,papadopulos2015}. The independent evolution of these phenomena in such taxonomically distant groups suggest that general evolutionary mechanisms may be involved, but inferring the causes of molecular evolution and patterns of polymorphism in genomic regions that lack recombination is a longstanding challenge for both theoreticians and experimentalists \citep{charlesworth1978,feldman1980evolution,barton1995general,charlesworth1996CB,otto1997deleterious,charlesworth2000degeneration,mcvean2000effects}.

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Thus, genes on the Y chromosome are expected to experience an effective population size ($N_{e}$) that is 1/4 that of autosomal genes, whereas the  $N_{e}$ for genes on the X chromosome should be 3/4 that of autosomal genes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that that the equilibrium level of neutral polymorphism - proportional to the product of $N_{e}$ and the neutral mutation rate, $\mu$ - should be lower for Y-linked genes than for their X-linked counterparts. In the absence of recombination, genes on the Y chromosome are also expected to be in strong linkage disequilibrium, making them vulnerable to diversity loss due to selection against strongly deleterious mutations (background selection) and selective sweeps of strongly beneficial mutations (genetic hitchhiking). Furthermore, the build-up of linkage disequilibrium between selected mutations on the Y means that selection will act non-independently across the chromosome, such that selection at a focal site may "interfere" with selection at sites with which it is linked \citep{hill1966HReffect}. A large body of work has now shown that in theory such "selective interference" effects can substantially reduce both the efficacy of selection and the level of neutral variability \citep{fisher1930genetical, muller1964relation, hill1966HReffect, mcvean2000,KaiserCharlesworth,good2014genetic}. These arguments all suggest that non-recombining Y chromosomes should harbor a lower amount of neutral genetic variability than predicted based on the number of Y chromosomes in a population, but the extent to which background selection, genetic hitchhiking, or selective interference have affected chromosome-wide levels of diversity in natural populations, and the relative importance of these processes, is not well-understood.

In addition to reduced diversity arising from selection, in species with female-biased sex ratios or extensive male-male competition, high variance in male reproductive success is also expected reduce the $N_{e}$ experience by genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting that inferences about the effects positive of purifying selection need to be distinguished from these neutral processes. Because variance in male reproductive success reduces both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this can therefore be obtained by comparing levels of neutral diversity on X and Y chromosomes relative to values on autosomes. For example, high variance in male reproductive success causes a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans, for example, have suggested that the inflated X/A ratio is due to a historical excess of breeding females compared to males \citep{hammer2008sex} (and see \citep{bustamante2009,hammer2010,cotter2016genetic}).

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on sex chromosomes \citep{ellegren2011,bachtrog2013NRG}, we know very little about the influence of sex ratio variation or linked selection in determining levels of diversity on more recently evolved sex chromosomes. The time scales over which these different effects are likely to be important is therefore not well understood. In humans, estimates of Y-linked diversity are considerably lower than predicted under neutral models, and simulations suggest that levels of diversity are consistent with strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA) \citep{lahn1999four,ross2005dna}, is not immediately obvious whether purifying selection might have such strong effects on Y chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last \textasciitilde 20 MYA in the case of plants \citep{charlesworth2015plant}). Y-linked diversity loss might on the one hand be expected to be lower in younger systems due to a shorter history of recombination suppression. On the other hand, simulations of strong selection models (background selection and genetic hitchhiking) suggest that these processes may have the greatest effects during the earliest stages of sex chromosome evolution, before the Y has lost many of its genes \citep{bachtrog2008temporal}. Moreover, there is clear evidence from simulation studies that even weak purifying selection, if acting non-independently across large number of sites, can generate substantial deviations from neutrality, whereas classic background selection theory breaks down in such cases \citep{mcvean2000, comeron2002population, KaiserCharlesworth, good2014genetic}. Given that a large number of sites are likely to be in linkage disequilibrium on a recently evolved Y chromosome, such "interference selection" \textit{sensu} \citep{good2014genetic} is a good candidate model for exploring the evolutionary dynamics of young plant sex chromosomes.

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution, we analyzed neutral polymorphism levels on X, Y, and autosomal chromosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with heteromorphic X and Y chromosomes that evolved \textasciitilde 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making Y chromosomes in this species over 100 million years younger than the highly degenerated Y chromosomes in mammals \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the occurrence of an interesting polymorphic in sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct populations ( so-called "chromosomal races") \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species is thought to have originated through an X-autosome fusion, with the XY system maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, there is evidence that both Y's have undergone gene loss and functional deterioration \citep{hough2014}. Here, to simplify our comparison of polymorphism levels on X, Y, and autosomes, we focus only on the ancestral Y chromosome, which occurs in both sex chromosome races.

Of particular importance for our study, \textit{R. hastatulus} populations have also been found to consistently exhibit female-biased reproductive sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Female-biased sex ratios are not uncommon in dioecious plants with heteromorphic X and Y chromosomes \citep{field2013comparative,hough2013evolutionarily}, and their occurrence in \textit{R. hastatulus} provides an excellent opportunity to investigate both the demographic and selective factors contributing to sex-linked variability.

@@

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether our estimated levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, we compared our estimated levels diversity to predictions from a model for the expected $N_{e}$ of males and females when the variance in offpring number is smaller or larger than the standardly assumed Poisson \citep{kimura1964number, hedrick2011genetics}. As we were primarily interested in the influence of variance in male reproductive success on Y-linked $N_{e}$, and the parameter space within which this effect might generate Y/A and X/A ratios consistent with our data, we ignored non-poission variance in female offspring numbers, and calculated male $N_{e}$ as:

@@

where $k_{m}$ and $V_{k_{m}}$ are the mean and variance in male offspring number, respectively, in a population of $N_{m}$ reproducing males, and $\overline{k}$ is the total mean number of progeny \citep{kimura1964number}. Given expected values of $N_e_{m}$ for a range of values for $V_{k_{m}}$ and $N_{m}$, the corresponding neutral expectation for autosomal and sex-linked $N_{e}$ at equilibrium were calculated as:

\begin{equation}
N_{e_{A}} = \frac{4N_{m}N_{f}}{N_{m}+N_{f}} \label{eq:Ne}
\end{equation}

\begin{equation}
N_{e_{X}} = \frac{9N_{m}N_{f}}{4N_{m}+2N_{f}} \label{eq:NeX}
\end{equation}

\begin{equation}
N_{e_{Y}} = \frac{N_{m}}{2} \label{eq:NeY}
\end{equation}

respectively, following \citep{wright1931evolution}. Note that with equal sex ratios $N_{f} = N_{m}$, and Poisson-distributed offspring numbers ($V_{k} = \overline{k}$), then $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$ and in the standard model. Figure 1 shows these predictions as a function of the sex ratio for a range values of $V_{k_{m}}$, and we tested the fit of our observed Y/A and X/A diversity ratios to model predictions (shown in Figure 2) assuming the empirically-estimated \textit{R. hastatulus} population sex ratio of 0.6 \citep{pickup2013influence}.

@@

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious mutations from our polymorphism data for X-linked genes using the method of \citep{keightley2007joint}, which fits a gamma distribution of selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphism. We then used this estimated gamma distribution to parameterize the simulations, initializing each with our estimated $\theta$ from autosomal genes, but appropriately adjusted to reflect the expected reduction in $N_{e}_{Y}$ for a sex ratio of $N_{f}/(N_{f}+N_{m})=0.6$. To match our sample size and the number of synonymous sites sample from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and contained 45,331bp of linked neutral sequence from which we calculated silent site diversity. To examine the expected reduction in diversity ($\pi/\pi_{0}$) as a function of the number of selected sites ($L$), we ran simulations over a range of values of $L$, up to a maximum of 5x$10^{6}$ (Figure 3A). To estimate $L$ from our data, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi__{simulated}$, was not significantly different from our empirical estimate, $\pi__{observed}$ (Figure 3B).

\section*{Results and Discussion}
\subsection*{Extensive loss of Y-chromosome diversity}
Our analysis of polymorphism levels across the genome in the dioecious plant \textit{R. hastatulus} reveals that neutral diversity on the Y chromosome is substantially lower than expected from a standard neutral model. In particular, neutral diversity on Y-linked genes $\theta_{Y}$ is on average \textasciitilde 2.1\% of the value for their homologues on the X chromosome. Talking 1/4 of the mean autosomal diversity as the equilibrium expectation under neutrality \citep{wright1931evolution}, this corresponds to a chromosome-wide reduction of \textasciitilde 93\%. Note that by comparing X and Y diversity to autosomal diversity, our results indicate that the difference between X and Y homologues is not due to an elevation of diversity on the X chromosome, but rather a Y-specific reduction. Our results do suggest a slight elevation in X-linked diversity relative to the neutral prediction of X/A=0.75, but this difference is not significantly different from the expectation for the empirically-estimated ratio of 0.6. Although our estimates of diversity have not beed normalized by divergence, previous work has shown that the average synonymous substitution rate, between \textit{R. hastatulus} and the non-dioecious outgroup \textit{R. bucephalophorus} is not significantly different for sex-linked (0.2016) or autosomal genes (0.219) \citep{hough2014}. It is therefore unlikely that our results are caused by mutation rate differences between sex-linked and autosomal genes.

\subsection*{Female biased sex ratios and variance in male offspring number}

%\begin{figure}[htbp]
%\centering
%\noindent
%\includegraphics[width=\linewidth]{figure2.jpg}
%\caption{new figure coming soon).
%}
%\label{fig:spectrum}
%\end{figure}
